---
title: "Less is More? Big Ten Teams Run Less Plays, But Powerhouses Remain Unaffected"
author: "Sashwat Venkatesh"
date: "2023-10-10"
categories: [news]
format:
  html:
    code-fold: true
---

The NCAA's new clock rules in football were implemented with a view to shorten games and make the viewer experience more palatable. Have they done that? Let's take a look within the context of the Big Ten:

```{r, output=false}
library(tidyverse)
```

```{r}
games2022 <- read_csv("https://dwillis.github.io/sports-data-files/footballlogs22.csv")
games2023 <- read_csv("https://dwillis.github.io/sports-data-files/footballlogs23.csv")

big10 <- c("Maryland Terrapins", "Illinois Fightning Illini", "Wisconsin Badgers", "Michigan State Spartans", "Michigan Wolverines", "Indiana Hoosiers", "Ohio State Buckeyes", "Penn State Nittany Lions", "Rutgers Scarlet Knights", "Iowa Hawkeyes", "Minnesota Golden Gophers", "Nebraska Cornhuskers", "Northwestern Wildcats", "Purdue Boilermakers")

b10_games2022 <- games2022 |> filter(TeamFull %in% big10)
b10_games2023 <- games2023 |> filter(TeamFull %in% big10)

games2022_playsPerGame <- b10_games2022 |> group_by(TeamFull) |> summarize(avgOffPlays = sum(OffensivePlays)/n())
games2023_playsPerGame <- b10_games2023 |> group_by(TeamFull) |> summarize(avgOffPlays = sum(OffensivePlays)/n())

ggplot() + 
  geom_bar(
    data=games2022_playsPerGame, 
    aes(x=TeamFull, weight=avgOffPlays)
  ) +
  ylab("Offensive Plays Per Game") +
  xlab("Team") +
  ggtitle("Offensive Plays Per Game by Big Ten Teams, 2022")+
  coord_flip()

```
Looking at the Big Ten teams in 2022, we can see that the perennial Big Ten powerhouses, like Ohio State and Penn State, weren't necessarily the ones running the most offensive plays. The bar chart we've generated here indicates that Northwestern and Purdue, teams that aren't known for "running" the Big ten like OSU and Penn State are, might suffer from some sort of offensive inefficiency.


```{r}
ggplot() + 
  geom_bar(
    data=games2023_playsPerGame, 
    aes(x=TeamFull, weight=avgOffPlays)
  ) +
  ylab("Offensive Plays Per Game") +
  xlab("Team") +
  ggtitle("Offensive Plays Per Game by Big Ten Teams, 2023")+
  coord_flip()
```
Once again, in 2023, Purdue is amongst the top teams in terms of Offensive Plays run per game, and sit at a measly 2-4 on the season.


```{r}
differentials <- data.frame(
  Team = games2022_playsPerGame$TeamFull,
  playsPerGame2023 = games2023_playsPerGame$avgOffPlays,
  playsPerGame2022 = games2022_playsPerGame$avgOffPlays
)
differentials <- differentials |> mutate(Differential = differentials$playsPerGame2023 - differentials$playsPerGame2022)

differentials |> arrange(desc(Differential))
```

Looking at the differentials between seasons, the new clock rule intended to shorten games seems to be doing just that -- in the Big Ten at least. Of the Big Ten's teams, just 3 have logged an increased in an average number of plays run per game. Those teams, Penn State, Wisconsin, and Michigan State, are 5-0, 4-1, and 2-3, respectively, indicating that there's something missing here. What are Penn State and Wisconsin doing differently to only have one loss between them?

(The Badgers' one loss was to #19 Washington State by 9 points, a result of falling behind early on in what was otherwise an even matchup on paper)