---
title: "Welcome Willard: A New Terrapin Identity"
author: "Sashwat Venkatesh"
date: "2023-11-14"
categories: [news]
format:
  html:
    code-fold: true
---

```{r}
suppressMessages(library(tidyverse))
suppressMessages(library(ggplot2))
suppressMessages(library(formattable))

suppressWarnings(suppressMessages(bigdf <- read_csv("https://thescoop.org/sports-data-files/cbblogs1524.csv")))
suppressMessages(sh <- bigdf |> filter(TeamFull == "Seton Hall Pirates") |> filter(Date < "2022-07-01"))
suppressMessages(md2 <- bigdf |> filter(TeamFull == "Maryland Terrapins") |> filter(Date > "2022-07-02"))
suppressMessages(md1 <- bigdf |> filter(TeamFull == "Maryland Terrapins") |> filter(Date < "2022-07-02"))
```

```{r}
t3 <- md1 |> group_by(Season) |>summarise(
  oppg = mean(OpponentScore),
  stlg = mean(TeamSteals),
  opptov = mean(OpponentTurnovers),
  fta = mean(TeamFTA),
  oppFTA = mean(OpponentFTA),
  oppORBg = mean(OpponentOffRebounds)
)
formattable(t3, list(Season = color_bar("red")))
```


```{r}
t1 <- sh |> group_by(Season) |>summarise(
  oppg = mean(OpponentScore),
  stlg = mean(TeamSteals),
  opptov = mean(OpponentTurnovers),
  fta = mean(TeamFTA),
  oppFTA = mean(OpponentFTA),
  oppORBg = mean(OpponentOffRebounds)
)
formattable(t1, list(Season = color_bar("blue")))
```

```{r}
suppressWarnings(t2 <- md2 |> group_by(Season) |>summarise(
  oppg = mean(OpponentScore),
  stlg = mean(TeamSteals),
  opptov = mean(OpponentTurnovers),
  fta = mean(TeamFTA),
  oppFTA = mean(OpponentFTA),
  oppORBg = mean(OpponentOffRebounds)
))

suppressWarnings(formattable(t2, list(Season = color_bar("red"))))
```

After firing head coach Mark Turgeon during the 2021-22 season, the Maryland Terrapins completed the season under interim coach Danny Manning, and on March 21, 2022, the Terrapins announced the hiring of Seton Hall head coach Kevin Willard to lead the program. In his first season, Willard would lead the Terrapins to a 8-0 start, and end the season with a perfect home record, along with an NCAA tournament appearance.

With Willard's hiring, Terrapin fans hoped that a change in identity after the school's first losing season in nearly 30 years would be near. Mark Turgeon, otherwise sometimes known as "Turgeon the Surgeon" during his playing days for his ability to "carve up" defenses as a player, brought this offensive mindset to his coaching philosophy. In his time leading the Terrapins, Turgeon was known for high-flying offenses.

Kevin Willard's Seton Hall teams, statistically, were solid defensively, taking the pressure off their offense. This defensive focus has made it to College Park, as we can see in the two above tables displaying Willard's coaching history.

Looking at defensive statistics, we can see that in 2022-23, the Terrapins did adopt Willard's defensive mentality, lowering their points allowed per game, as well as increasing steals and opponent turnovers. Unfortunately, on paper, while the Terrapins look strong defensively, their offense has not been up to par, dropping the Terps to a shaky 1-3 start this season. 

While the old adage says defense wins championships, perhaps the Terrapins and Kevin Willard need to find a way to complement their strong defense with an equal or at least close-enough offense.